---
title: "How to compare models with survxai"
author: "Alicja Gosiewska, Aleksandra Grudziaz"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{How to compare models with survxai}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```



# Introduction
Package *survxai* contains functions for creating a unified representation of a survival models. Such representations can be further processed by various survival explainers. Tools implemented in *survxai* help to understand how input variables are used in the model and what impact do they have on final model prediction.

The analyses carried out using this package can be divided into two parts: local analyses of new observations and global analyses showing the structures of survival models. In this vignette we present both, global and local explanations. 

Methods and functions in  *survxai* package are based on [*DALEX* package](https://github.com/pbiecek/DALEX). 

# Use case - data

## Data set
We use the data from the Mayo Clinic trial in primary biliary cirrhosis (PBC) of the liver conducted between 1974 and 1984. A total of 424 PBC patients, referred to Mayo Clinic during that ten-year interval, met eligibility criteria for the randomized placebo controlled trial of the drug D-penicillamine. First 312 cases in the data set participated in the randomized trial and contain largely complete data.
The `pbc` data is included in the [*randomForestSRC* package](https://cran.r-project.org/web/packages/randomForestSRC/index.html).

We base models on three variables from `pbc` data set. They are `stage`, `sex`, and `bili`. For this purpouse we create a new data set `pbc_smaller`. We also change `days` for `years`.

```{r dataset}
data(pbc, package = "randomForestSRC")
pbc <- pbc[complete.cases(pbc),]
pbc_smaller <- pbc[,c("days", "status","sex", "bili", "stage")]
pbc_smaller$years <- pbc$days/365
pbc_smaller <- pbc_smaller[,-1]
head(pbc_smaller)
```

We also convert variables `sex` and `stage` to factors.

```{r}
pbc_smaller$sex <- as.factor(pbc_smaller$sex)
pbc_smaller$stage <- as.factor(pbc_smaller$stage)
```

We divide our data set into train and test sets.
```{r}
smp_size <- floor(0.5 * nrow(pbc_smaller))
set.seed(123)
train_ind <- sample(seq_len(nrow(pbc_smaller)), size = smp_size)

train <- pbc_smaller[train_ind, ]
test <- pbc_smaller[-train_ind, ]
```


## Models

In this section, we create 3 models: Cox Proportional Hazards Model, Random Forests for Survival, and Parametric Survival Regression Model.


### Cox Proportional Hazards Model

Cox proportional hazards model is based on five variables: `age`, `treatment`, `stage`, `sex`, and `bili`.

```{r, models}
library(rms)

cph_model <- cph(Surv(years, status)~., data = train, surv = TRUE, x = TRUE, y=TRUE)
```

### Random Forests for Survival

Random Forests for Survival is based on five variables: `age`, `treatment`, `stage`, `sex`, and `bili`.

```{r}
library(randomForestSRC)
set.seed(1994)
rf_model <- rfsrc(Surv(years, status)~., data = train)
```

### Parametric Survival Regression Model

Parametric Survival Regression Model is based on five variables: `age`, `treatment`, `stage`, `sex`, and `bili`.

```{r}
library(survival)
reg_model <- survreg(Surv(years, status)~., data = train, x = TRUE)
```

# Explanation

First, we have to create survival explainers - objects that wrap-up the black-box model with meta-data. Explainers unify model interfacing.

Some models require custom predict function. Examples are in [Custom predict function for survival models vignette](https://mi2datalab.github.io/survxai/articles/Custom_predict_for_survival_models.html).

Predict functions for `cph` and `rfsrc` objects are implemented.
```{r}
library(survxai)

surve_cph <- explain(model = cph_model,
                     data = test[,-c(1,5)], 
                     y = Surv(test$years, test$status))

print(surve_cph)
surve_rf <- explain(model = rf_model, 
                     label = "random forest",
                     data = test[,-c(1,5)], 
                     y = Surv(test$years, test$status))

print(surve_rf)
```

For model created by `survreg()` function we have to add specific custom predict function. 
We add this function to explainer by `predict_function` parameter.

```{r}
library(CFC)

custom_predict <- function(model, newdata, times){
  times <- sort(times)
  vars <- all.vars(model$call[[2]][[2]])
  n_vars <- which(colnames(newdata) %in% vars)
  if(length(n_vars)>0){
    newdata <- newdata[,-c(n_vars)]
  }
  model$x <- model.matrix(~., newdata)
  res <- matrix(ncol = length(times), nrow = nrow(newdata))
  for(i in 1:nrow(newdata)) {
    res[i,] <- cfc.survreg.survprob(t = times, args = model, n = i)    
  }
  return(res)
}

surve_reg <- explain(model = reg_model,
                     data = test[,-c(1,5)], 
                     y = Surv(test$years, test$status),
                     predict_function = custom_predict)
print(surve_reg)
```


## Global explanations
At the beginning, we are dealing with global explanations. We focus on explanations of the global and conditional model structure.

### Model performance

Currently, in the *survxai* package is implemented only the `BS` type of model performance.
In this metod for each time point we compute the prediction error for model.
We compute `model_performance` explainers for all three models created above.

```{r}
mp_cph <- model_performance(surve_cph)
mp_rf <- model_performance(surve_rf)
mp_reg <- model_performance(surve_reg)
```

For each explainer we can print values using generic `print()` function.
```{r}
print(mp_cph)
```

We plot all three models using generic `plot()` function.
```{r}
plot(mp_cph, mp_rf, mp_reg) + ylim(c(0,0.30))
```

On the plot, we see that the Cox Proportional Hazards Model and Parametric Survival Regression Model have simmilar performance - prediction errors for these models are almost the same. For the Survival Random Forest Model we have the smallest prediction errors.

### Variable response
Variable response explainers are designed to better understand the relation between a variable and a model output.
We compute `variable_response` explainers for all three models created above.

#### Factor variable
First, we compute explainers for the factor variable `sex`.
```{r}
vr_cph_sex <- variable_response(surve_cph, "sex")
vr_rf_sex <- variable_response(surve_rf, "sex")
vr_reg_sex <- variable_response(surve_reg, "sex")
```

For each explainer we can print values using generic `print()` function.
```{r}
print(vr_cph_sex)
```

We visualize explainers on one plot.
```{r, fig.height=10}
plot(vr_cph_sex, vr_rf_sex, vr_reg_sex)
```

We see that in all models patients with `sex=1` have higher probability of survival.

In `survival_variable_response` plots we can also compare models in division into all variable levels using parameter `split = "variable"`.

```{r}
plot(vr_cph_sex, vr_rf_sex, vr_reg_sex, split = "variable")

```


#### Continuous variable
We can also compute the `pdp` plots for numeric variables.
For this, we divide values of variable `bili` into 6 intervals.
```{r}
vr_cph_bili <- variable_response(surve_cph, "bili")
vr_rf_bili <- variable_response(surve_rf, "bili")
vr_reg_bili <- variable_response(surve_reg, "bili")
```

```{r, fig.height=10}
plot(vr_cph_bili, vr_rf_bili, vr_reg_bili)
```

As mentioned above we can split our comparison by each variable level.
```{r, fig.height=10}
plot(vr_cph_bili, vr_rf_bili, vr_reg_bili, split = "variable")
```


## Local explanations
The next aspects of survival model analysis are local explanations - the explanations for one, new observation.

```{r}
single_observation <- test[1,-c(1,5)]
single_observation
```


### Ceteris paribus
Ceteris Paribus Plots for survival models are survival curves around one observation. Each curve represent observation with different value of chosen variable. For factor variables curves covers all possible values, for numeric variables values are divided into quantiles.
```{r}
cp_cph <- ceteris_paribus(surve_cph, single_observation)
cp_rf <- ceteris_paribus(surve_rf, single_observation)
cp_reg <- ceteris_paribus(surve_reg, single_observation)
```

For each ceteris paribus explainer we can print values using generic `print()` function.
```{r}
print(cp_cph)
```

```{r, fig.height=10}
plot(cp_cph, scale_type = "gradient", scale_col = c("red", "blue"))

plot(cp_rf, scale_type = "gradient", scale_col = c("red", "blue"))
plot(cp_reg, scale_type = "gradient", scale_col = c("red", "blue"))
```

On the plots above we see that for variables `bili` and `sex` `coxph` and `survreg` model have simmilar performance, but `randomForest` model has different predictions of survival probability.

### Prediction breakdown
Break Down Plots for survival models compare, by default, differences in predictions for median of time. We can change the time of computing the contributions for each variable using `time` parameter. 
```{r}
broken_prediction_cph <- prediction_breakdown(surve_cph, single_observation)
broken_prediction_rf <- prediction_breakdown(surve_rf, single_observation)
broken_prediction_reg <- prediction_breakdown(surve_reg, single_observation)
```

For each prediction breakdown explainer we can print values using generic `print()` function.
```{r}
print(broken_prediction_cph)
```


```{r}
plot(broken_prediction_cph, scale_col = c("red", "blue"), lines_type = 2)
plot(broken_prediction_rf, scale_col = c("red", "blue"), lines_type = 2)
plot(broken_prediction_reg, scale_col = c("red", "blue"), lines_type = 2)
```

This plots helps to understand the factors that drive survival probability for a single observation.

For all models variable `bili` is the variable with the largest contribution. But the second variable is different randomForest (`age`) and coxph, survreg (`stage`). 

We can also compute the variables contribution for chosen survival probability value.
```{r}
broken_prediction_cph <- prediction_breakdown(surve_cph, single_observation, prob = 0.8)
broken_prediction_rf <- prediction_breakdown(surve_rf, single_observation, prob = 0.8)
broken_prediction_reg <- prediction_breakdown(surve_reg, single_observation, prob = 0.8)
```

```{r}
plot(broken_prediction_cph, scale_col = c("red", "blue"), lines_type = 2)
plot(broken_prediction_rf, scale_col = c("red", "blue"), lines_type = 2)
plot(broken_prediction_reg, scale_col = c("red", "blue"), lines_type = 2)
```

We can compare this models on one plot.

```{r, fig.height=10}
plot(broken_prediction_cph, broken_prediction_rf, broken_prediction_reg, scale_col = c("red", "blue"), lines_type = 2)
```


For the transparency of the charts we recommend drawning them separately.
